---
title: "Class 9: Candy Mini-Project"
author: "Linh Tran (PID:A16435846)"
format: pdf
---
For this mini-project, we will explore FiveThirtyEight's Halloween Candy dataset (538) to determine the top and worst ranked candies based on different factors such as price and the types.

# Importing Data from **FiveThirtyEight GitHub repo**

First, we have to import candy data by downloading `candy-data.csv` and using `read.csv()`.

```{r}
candy_file <- "candy-data.txt"
candy<-read.csv(candy_file, row.names=1)
head(candy)
```

> Q1. How many different candy types are in this dataset?

```{r}
nrow(candy)
```

85 candy types

> Q2. How many fruity candy types are in the dataset?

```{r}
sum(candy$fruity)
```

38 fruity candy types

# What is your favorite candy?

> Q3. What is your favorite candy in the dataset and what is it’s winpercent value?

We can look at `winpercent` to find the percentage of people who prefer a particular candy over another randomly chosen candy.

My favorite candy is Kit Kat and it's winpercent value is 47.83

```{r}
candy["Whoppers", ]$winpercent
```

> Q4. What is the winpercent value for “Kit Kat”?

```{r}
candy["Kit Kat", ]$winpercent
```

> Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy[ "Tootsie Roll Snack Bars", ]$winpercent
```

A useful function from "skimr" that can help giving a quick overview of a dataset is `skim()` first, we have to download `skimr` with `install.packages("skimr")`

```{r}
library("skimr")
skim(candy)
```

> Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

The variables that look different from the majority are "s**ugarpercent", "pricepercent",** and **"winpercent".**

> Q7. What do you think a zero and one represent for the candy\$chocolate column?

I believe a zero shows that the corresponding candy type based on the row is not chocolate whereas a one would show that it is chocolate.

> Q8. Plot a histogram of winpercent values

We can plot a histogram using either `hist()` or `ggplot()`

```{r}
library("ggplot2")

ggplot(candy)+
  aes( x=candy$winpercent)+
  geom_histogram()
```

```{r}
hist(candy$winpercent)
```

> Q9. Is the distribution of winpercent values symmetrical?

The distribution of winpercent values is assymetrical, it is right skewed.

> Q10. Is the center of the distribution above or below 50%?

The center of the distribution is below 50%

> Q11. On average is chocolate candy higher or lower ranked than fruit candy?

Using `as.logical()`, we can turn 1's and 0's into logical and assign them to their corresponding category. Furthermore, we can take the average with `mean()`

```{r}
mean(candy$winpercent[as.logical(candy$chocolate)])

mean(candy$winpercent[as.logical(candy$fruity)])

```

From the results, we see that chocolate has a higher winpercent value on average than fruit candy; therefore, chocolate ranks higher

> Q12. Is this difference statistically significant?

The t-test can be performed with `t.test()`

```{r}
t.test(candy$winpercent[as.logical(candy$chocolate)],
       candy$winpercent[as.logical(candy$fruity)])

```

The t-test shows that there is a significant difference in the menas of the two candies, that is chocolate is likely ranked higher.

# Overall Candy Rankings

> Q13. What are the five least liked candy types in this set?

we can use either bse R with `head()` or `dplyr` to arrange the candy types.

```{r}
head(candy[order(candy$winpercent),], n=5)

```

```{r}
library(dplyr)
candy %>% arrange(winpercent) %>% head(5)

```

Both methods show that the least liked 5 are Nik L Nip, Boston Baked Beans, Chiclets, Super Bubble, and Jawbusters

> Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
tail(candy[order(candy$winpercent),], n=5)

```

```{r}
candy %>% arrange(winpercent) %>% tail(5)

```

Both methods showed that the top 5 are Snickers, Kit Kat, Twix, Reese's Miniatures, and Reese's Peanut Butter cup.

I like dplyr more, despite having to first retrieve it with `library(),` because the code is more straight forward.

> Q15. Make a first barplot of candy ranking based on winpercent values.

```{r}
ggplot(candy) + 
  aes(winpercent, rownames(candy)) +
  geom_col()
```

> Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy), winpercent)) +
  geom_col()
```

We can also add color

```{r}
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"

ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols) 
```

> Q17. What is the worst ranked chocolate candy?

The worst ranked chocolate candy is Sixlets.

> Q18. What is the best ranked fruity candy?

The best ranked fruity candy is Starburst.

# Taking a look at "*pricepercent"*

> Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

To determine the highest ranked candy in terms of **winpercent** for the least money, we would have to plot "**winpercent"** vs "**pricepercent"** using `geom_label()` and `geom_text_repel()` from **ggrepel**

```{r}
library(ggrepel)
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "violet"


ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 5)
```
Based on the plot, I believe the highest ranked for the least money is Reeses Miniatures, which has a winpercent value of more than 80 and pricepercent value of slightly above 0.25.

> Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

We can use `order()` to arrange the candies based on **pricepercent** 

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )
```

The most expensive candies are in the order of  Nik L Nip, Nestle Smarties, Ring pop, Hersheys Krackel, Hersheys Milk Chocolate. Based on the result from **Q13.**, Nik L Nip is also the least popular 

> Q21. Make a barplot again with geom_col() this time using pricepercent and then improve this step by step, first ordering the x-axis by value and finally making a so called “dot chat” or “lollipop” chart by swapping geom_col() for geom_point() + geom_segment().

```{r}
ggplot(candy) +
  aes(pricepercent, reorder(rownames(candy), pricepercent)) +
  geom_segment(aes(yend = reorder(rownames(candy), pricepercent), 
                   xend = 0), col="gray40") +
    geom_point()
```


# Exploring the correlation structure
We can view how varianles interact by looking at correlation plots, which can be made with `corrplot()`

> Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

```{r}
library(corrplot)

cij <- cor(candy)
corrplot(cij)
```
The two variables that are anti-correlated are chocolate and fruity candy

> Q23. Similarly, what two variables are most positively correlated?

The two variables that are most positively correlated are bar candy and chocolate

# Principal Component Analysis
Lastly, we can apply PCA with `prcom()` and plot PC1 vs PC2

```{r}
pca <- prcomp(candy, scale=TRUE)
summary(pca)
```
```{r}
plot(pca$x[,1:2])
```
We can cahnge the plot character and add color
```{r}
plot(pca$x[,1:2], col=my_cols, pch=16)
```

Using the PCA results and Candy data, we can make a new data-frame with `cbind()`

```{r}
my_data <- cbind(candy, pca$x[,1:3])

p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols)

p
```
We then add title and substitle, as well as further label the plot with **ggrepel**. An interactive plot can also be created with 

```{r}
p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 7)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```

(I was unable to install ggplotly)


```{r}
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```

> Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

THe variables picked up strongly by PC1 in the positive direction are fruit, hard, and pluribus. They make sense because it they seem to correlate and in real life I also associate them with one another even without looking at graphs.
