---
title: "Class 8 Mini project"
author: "Linh Tran (PID:A16435846)"
format: pdf
---

## About
In today's lab, we will work with fine needle aspiration (FNA) of breast mass data from the University of Wisconsin.


## Data Import

```{r}
wisc.df <- read.csv("WisconsinCancer.csv", row.names=1)
head(wisc.df)
```
>Q1. How many patients/individuals/samples are in this dataset

```{r}
nrow(wisc.df)
```

>Q2. How many of the observations have a malignant diagnosis?

Use `table()` function
```{r}
table(wisc.df$diagnosis)
```

This can also be used

```{r}
sum(wisc.df$diagnosis == "M")

```

```{r}
colnames(wisc.df)
```




>Q3. How many variables/features in the data are suffixed with _mean?

```{r}
ncol(wisc.df)
```
```{r}
grep("_mean", colnames(wisc.df), value=T)
```


```{r}
inds <- grep("_mean", colnames(wisc.df))
length(inds)
```


## Initial Analysis
Befoe analysis, I want to take out the expert diagnoses column(a.k.a. the answer) from out dataset

```{r}
diagnosis <- as.factor(wisc.df$diagnosis)
diagnosis
```


```{r}
wisc.data <- wisc.df[,-1]
```




## Clustering

We can try a kmean() clustering first

```{r}
km <- kmeans(wisc.data, centers=2)

```

```{r}
table(km$cluster)
```

Cross-table

```{r}
table(km$cluster, diagnosis)
```
Let's try `hclust()` the key input required for `hclust()` is a distance matrix as produced by the `dist()` function.
```{r}
hc <- hclust(dist(wisc.data))
```

```{r}
plot( hc)
```


## PCA
Do we need to scale the data?

We can loook at the sd of each column

```{r}
round(apply(wisc.data,2,sd))
```

Yes we need ot scale. We will run `pcomp()` with `scale=TRUE`.

```{r}
wisc.pr <- prcomp(wisc.data, scale=TRUE)
summary(wisc.pr)
```

Generate our main PCA plot (score plot, PC1 vs PC2 plot)...

```{r}
library(ggplot2)

res <- as.data.frame(wisc.pr$x)
```

```{r}
ggplot(res) +
  aes(x=PC1, y=PC2, col=diagnosis)+
  geom_point()
```

```{r}
ggplot(res) +
  aes(x=PC1, y=PC4, col=diagnosis)+
  geom_point()
```


## Combining methods

Clustering on PCA results
Using the minimum number of principal components required to describe at least 90% of the variability in the data, create a hierarchical clustering model with the linkage method="ward.D2". We use Ward’s criterion here because it is based on multidimensional variance like principal components analysis. Assign the results to wisc.pr.hclust.

```{r}
d <- dist(wisc.pr$x[,1:3])
hc <- hclust(d, method="ward.D2")
plot(hc)
```
To get my clustering result/membership vector, I need to "cut" the tree with the `cutree()` function
```{r}
grps <- cutree(hc, k=2)
```

>Q. How many patients are in each cluster group?

```{r}
table(grps)
```

```{r}
plot(res$PC1, res$PC2, col=grps)
```

## Prediction
We can use our PCA result (model) to do predictions, that is take new unseen data and project it onto our new PC variables

```{r}
#url <- "new_samples.csv"
url <- "https://tinyurl.com/new-samples-CSV"
new <- read.csv(url)
npc <- predict(wisc.pr, newdata=new)
npc
```

```{r}
plot(res$PC1, res$PC2, col=grps)
points(npc[,1], npc[,2], col="blue", pch=16, cex=3)
text(npc[,1], npc[,2], labels=c(1,2), col="white")
```

# Summary

Principal Component Analysis (PCA) is a super useful method for analyzing large datasets. It works by finding new variables (PCs) that capture the most variance from the original variables in your dataset.





